Variants of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus

### 2016-04-02 Papers/InProgress/NaturalDeduction/GodProof-ND.tex
Bjordal proof corollary added
Anderson proof: axiom 4 added

The aim of this paper is to present two detailed formalizations of ontological arguments in a natural deduction calculus. For a comprehensive introduction to natural deduction, the reader can consult \citep{prawitz}. The higher-order modal natural deduction calculus proposed here has special introduction and elimination rules for modalities, as defined in Section \ref{sec:calculus}, and is sound and complete relative to an axiomatic modal calculus, as proven in Section \ref{sec:compl-sound}.

@@

In G\"odel's ontological proof, we are proving a restricted modal collapse, which applies to one specific formula, the existence of a God-like being. The interest in the proof naturally decreases if a consequence of the axiomatization is a modal collapse for \emph{all} formulas. Therefore, an improvement would be obtained if the modal collapse was limited only to the property of being god-like or at least to a restricted collection of properties. Several solutions to the problem of modal collapse have been proposed.
Fitting, for instance, has argued that greater care has to be taken with the semantics of higher-order modal logics. Quantified variables may be rigid or flexible; and properties may be treated as intensional or extensional. Making the right choices may prevent the modal collapse \citep{fitting}[Sections 11.9 and 11.10]. Koon \citep{koon}, Anderson \citep{and90}[p. 292] and Sobel \citep{sobel2}[p. 133] also discuss the idea that the notion of property over which quantification is allowed might be too general and restrictions might be appropriate. Two alternative and more conservative solutions, requiring only modifications of the axioms and definitions but no change in the logic, are discussed in greater detail in the next section.

@@

ToDo: Proof of $\star_7$ is missing in the proof of lemma 1 below.

@@

\begin{scriptsize}
\begin{prooftree}
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$}  \dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$\forall \varphi. [P(\lambda x.\varphi(x)) \biimp \nec \varphi(\gamma)]$}\RightLabel{$\forall_E$}
\UIC{$P(\lambda x.\rho(x)) \biimp \nec \rho(\gamma)$}
\AXC{$ $} \RightLabel{5}
\UIC{$\nec \rho(\gamma)$} \RightLabel{$\imp_E$}
\BIC{$P(\lambda x.\rho(x))$} 
\AXC{$ $} \RightLabel{6}
\UIC{$G(y) $}\dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$\forall \varphi. [P(\lambda x.\varphi(x)) \biimp \nec \varphi(y)]$}
\UIC{$[P(\lambda x.\rho(x)) \biimp \nec \rho(y)]$}
\BIC{$\nec \rho(y)$}\RightLabel{$\nec_E$}
\UIC{$\omega':  \rho(y)$}\RightLabel{$\imp_I^6$}
\UIC{$\omega':  G(y)\imp \rho(y)$}\RightLabel{$\forall_I$}
\UIC{$\omega':  \forall y. (G(y)\imp \rho(y))$}\RightLabel{$\nec_I$}
\UIC{$\nec \forall y. (G(y)\imp \rho(y))$}\RightLabel{$\imp_I^5$}
\UIC{$[\subproof_7] \quad \nec \rho(\gamma)\imp \nec \forall y. (G(y)\imp \rho(y))$}
\end{prooftree}
\end{scriptsize}

%implication for equivalence in essence: 
\begin{small}
\begin{prooftree}
\AXC{Axiom \ref{A:A3}} 
\UIC{$P(G)$}
\AXC{$ $} \RightLabel{7}
\UIC{$\nec  \forall y. [G(y)\imp \varphi(y)]$}\RightLabel{$\wedge_I$}
\BIC{$P(G)\wedge \nec  \forall y. [G(y)\imp \varphi(y)]$}\RightLabel{$\imp_E$}
\AXC{Axiom \ref{A:A2}} 
\BIC{$P(\varphi)$}
\AXC{$\subproof_6$} \dashedLine 
\UIC{$  G(\gamma)$}  \RightLabel{D\ref{A:D1}}
\UIC{$ \all \varphi. [P(\varphi)\biimp \nec \varphi(\gamma)]$}  \RightLabel{$\all_E$}
\UIC{$ P(\varphi)\biimp \nec \varphi(\gamma)$}  \RightLabel{$\biimp_E$}
\BIC{$ \nec \varphi(\gamma)$} \RightLabel{$\nec_I$}
\UIC{$[\subproof_8] \quad \nec  \forall y. [G(y)\imp \varphi(y)]\imp \nec \varphi(\gamma)$}
\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$\subproof_7$ and $\subproof_8$} \dashedLine
\UIC{$\nec \varphi(\gamma)\biimp \nec  \forall y. [G(y)\imp \varphi(y)] $}
\UIC{$\forall \varphi. (\nec \varphi(\gamma)\biimp \nec  \forall y. [G(y)\imp \varphi(y)]) $}
\AXC{$ $} \RightLabel{6}
\UIC{$G(x)$}\RightLabel{$\wedge_I$}
\BIC{$\ess{G}{\gamma}$}
\AXC{Axiom \ref{A5}} \dashedLine
\UIC{$P(E)$}\dottedLine 
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$}\dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$\forall \varphi.(P(\varphi) \biimp \nec \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(E) \biimp \nec E(\gamma)$}\RightLabel{$\imp_E$}
\BIC{$\nec E(\gamma)$}
    \AXC{Axiom T} \dashedLine
    \UIC{$\nec E(\gamma) \imp E(\gamma)$} \RightLabel{$\imp_E$}
\BIC{$ E(\gamma)$} \dottedLine \RightLabel{D\ref{A:D3}}
\UIC{$ \all \varphi.[\ess{\varphi}{\gamma} \imp \nec \ex x.\varphi(x)] $}\RightLabel{$\all_E$}
\UIC{$ \ess{G}{\gamma} \imp \nec \ex x. G(x) $}\RightLabel{$\imp_E$}
\BIC{$\nec \ex x. G(x)$} \RightLabel{$\imp_I^1$}
\UIC{$\ex z. G(z) \imp \nec \ex x. G(x)$}
\end{prooftree}
\end{small}
\end{proof}

@@

@@

\begin{prooftree}
\AXC{$ $} \RightLabel{2}
\UIC{$\nec \neg \ex x.G_B(x) $} \RightLabel{$\nec_E$}
\UIC{$\omega: \neg \ex x.G_B(x) $} 
\AXC{$ $} \RightLabel{$1$}
\UIC{$\omega: G_B(x)$} \RightLabel{$\ex_I$}
\UIC{$\omega: \ex x.G_B(x) $} \RightLabel{$\imp_E$}
\BIC{$\omega:  \bot $}\RightLabel{$\imp_I^1$}
\UIC{$\omega:  \neg G_B(x) $} \RightLabel{$\imp_I$}
\UIC{$\omega:  G_B(x) \imp \neg G_B(x) $} \RightLabel{$\all_I$}
\UIC{$\omega:  \all x.(G_B(x) \imp \neg G_B(x)) $}  \RightLabel{$\nec_I$}
\UIC{$ \nec \all x.(G_B(x) \imp \neg G_B(x)) $}\RightLabel{$\wedge_I$}
\UIC{$P(\neg G_B)$}
\AXC{Axiom 1} \dashedLine
\UIC{$\all \varphi. [P(\neg G_B) \imp \neg P(G_B)]$} \RightLabel{$\all_E$}
\UIC{$ P( \neg G_B(x)) \imp \neg P(G_B) $}  \RightLabel{$\imp_E$}
\BIC{$\neg P(G_B) $} 
\AXC{$ $} \RightLabel{3} 
\UIC{$P(G_B)$} 
\BIC{$\bot$} \RightLabel{$\imp_I^2$}
\UIC{$ \neg \nec \neg \ex x.G_B(x)$} \doubleLine %\RightLabel{definition of $\pos$}
\UIC{$ \pos \ex x.G_B(x)$}  \RightLabel{$\imp_I^3$}
\UIC{$ P(G_B) \imp \pos \ex x.G_B(x) $}

@@

ToDo: State corollary and prove it, instead of making a forward reference to lemma L1. Then use the corollary in lemma 1.

Corollary 1 of section \ref{sec:newproof} follows by proving the positivity of godlikeness as in lemma \ref{B:L1} below.

@@

\begin{small}
\begin{prooftree}
\AXC{Axiom \ref{A:A5'}}\dashedLine
\UIC{$P_B (NE_B)$} \RightLabel{ D\ref{B:D1}}\dottedLine
\UIC{$\nec \all x. [G_B(x) \imp NE_B(x)]$}\RightLabel{ D\ref{B:D3}}\dottedLine
\UIC{$\nec \all x. [G_B(x) \imp \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$}  \RightLabel{ $\nec_E$}
\UIC{$\all x. [G_B(x) \imp \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$}  \RightLabel{ $\forall_E$}
\UIC{$ [G_B(x) \imp \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$}
\AXC{$ $} \RightLabel{ $2$}
\UIC{$G_B(x) $}  \RightLabel{ $\imp_E$}
\BIC{$\all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$} \RightLabel{ $\forall_E$}
\UIC{$(MCP(G_B, x) \imp \nec \ex y.G_B(y))]$}
\AXC{$ $}\dashedLine
\UIC{$MCP(G_B, x) $} \RightLabel{ $\imp_E$}
\BIC{$\nec \ex y.G_B(y)$}\RightLabel{ $\imp_I^2$}
\UIC{$G_B(x) \imp \nec \ex y.G_B(y)$}
\end{prooftree}
\end{small}
\end{proof}

>>>>

The aim of this paper is to present detailed formalizations of variants of ontological arguments in a natural deduction calculus. For a comprehensive introduction to natural deduction, the reader can consult \citep{prawitz}. The higher-order modal natural deduction calculus proposed here has special introduction and elimination rules for modalities, as defined in Section \ref{sec:calculus}, and is sound and complete relative to an axiomatic modal calculus, as proven in Section \ref{sec:compl-sound}.

@@

In G\"odel's ontological proof, we are proving a restricted modal collapse, which applies to one specific formula, the existence of a God-like being. The interest in the proof naturally decreases if a consequence of the axiomatization is a modal collapse for \emph{all} formulas. Therefore, an improvement would be obtained if the modal collapse was limited only to the property of being god-like or at least to a restricted collection of properties. Several solutions to the problem of modal collapse have been proposed.
Fitting, for instance, has argued that greater care has to be taken with the semantics of higher-order modal logics. Quantified variables may be rigid or flexible; and properties may be treated as intensional or extensional. Making the right choices may prevent the modal collapse \citep{fitting}[Sections 11.9 and 11.10]. Koon \citep{koon}, Anderson \citep{and90}[p. 292] and Sobel \citep{sobel2}[p. 133] also discuss the idea that the notion of property over which quantification is allowed might be too general and restrictions might be appropriate. Two alternative and more conservative solutions, requiring only modifications of the axioms and definitions but no change in the logic, are discussed in greater detail in the next sections.

@@

@@

\begin{scriptsize}
\begin{prooftree}
\AXC{$ $} \RightLabel{5}
\UIC{$\nec P(\rho)$} \RightLabel{$\nec_E$}
\UIC{$\omega: P(\rho)$}
\AXC{$ $} \RightLabel{4}
\UIC{$\omega: G(y)$}  \dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$\omega: \forall \varphi. [P(\lambda x.\varphi(x)) \biimp \nec \varphi(\gamma)]$}\RightLabel{$\forall_E$}
\UIC{$\omega: P(\rho) \biimp \nec \rho(y)$}\RightLabel{$\biimp_E$}
\BIC{$\omega: \nec \rho(y)$}\RightLabel{$\nec_E$}
\UIC{$\omega: \rho(y)$} \RightLabel{$\imp_I^4$}
\UIC{$\omega: G(y) \imp \rho(y)$}\RightLabel{$\forall_I$}
\UIC{$\omega: \forall y.(G(y) \imp \rho(y))$}\RightLabel{$\nec_I$}
\UIC{$\nec \forall y.(G(y) \imp \rho(y))$} \RightLabel{$\imp_I^5$}
\UIC{$\nec P(\rho)\imp \nec \forall y .(G(y)\imp \rho(y))$} \RightLabel{$\imp_E$}

\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$} \RightLabel{}\dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$\forall \varphi. [P(\lambda x.\varphi(x)) \biimp \nec \varphi(\gamma)]$}\RightLabel{$\forall_E$}
\UIC{$P(\lambda x.\rho(x)) \biimp \nec \rho(\gamma)$}
\AXC{$ $} \RightLabel{6}
\UIC{$\nec\rho(\gamma)$} \RightLabel{}
\BIC{$P(\rho)$} \RightLabel{$\imp_E$}

\AXC{Axiom \ref{A:A4}} \dashedLine
\UIC{$P(\rho)\imp \nec P(\rho)$} \RightLabel{$\imp_E$}
\BIC{$\nec P(\rho)$} \RightLabel{$\imp_E$}
\BIC{$\nec \forall y .(G(y)\imp \rho(y))$}\RightLabel{$\imp_I^6$}
\UIC{$[\subproof_7] \qquad \nec\rho(\gamma) \imp \nec \forall y.(G(y) \imp \rho(y))$}\RightLabel{$\forall_I$}
\end{prooftree}
\end{scriptsize}

%implication for equivalence in essence: 
\begin{small}
\begin{prooftree}
\AXC{Axiom \ref{A:A3}} 
\UIC{$P(G)$}
\AXC{$ $} \RightLabel{7}
\UIC{$\nec  \forall y. [G(y)\imp \varphi(y)]$}\RightLabel{$\wedge_I$}
\BIC{$P(G)\wedge \nec  \forall y. [G(y)\imp \varphi(y)]$}\RightLabel{$\imp_E$}
\AXC{Axiom \ref{A:A2}} 
\BIC{$P(\varphi)$}
\AXC{$\subproof_6$} \dashedLine 
\UIC{$  G(\gamma)$}  \RightLabel{D\ref{A:D1}}
\UIC{$ \all \varphi. [P(\varphi)\biimp \nec \varphi(\gamma)]$}  \RightLabel{$\all_E$}
\UIC{$ P(\varphi)\biimp \nec \varphi(\gamma)$}  \RightLabel{$\biimp_E$}
\BIC{$ \nec \varphi(\gamma)$} \RightLabel{$\imp_I^7$}
\UIC{$[\subproof_8] \quad \nec  \forall y. [G(y)\imp \varphi(y)]\imp \nec \varphi(\gamma)$}
\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$\subproof_7$ and $\subproof_8$} \dashedLine
\UIC{$\nec \varphi(\gamma)\biimp \nec  \forall y. [G(y)\imp \varphi(y)] $}
\UIC{$\forall \varphi. (\nec \varphi(\gamma)\biimp \nec  \forall y. [G(y)\imp \varphi(y)]) $}
\AXC{$ $} \RightLabel{6}
\UIC{$G(x)$}\RightLabel{$\wedge_I$}
\BIC{$\ess{G}{\gamma}$}
\AXC{Axiom \ref{A5}} \dashedLine
\UIC{$P(E)$}\dottedLine 
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$}\dottedLine \RightLabel{D\ref{A:D1}}
\UIC{$\forall \varphi.(P(\varphi) \biimp \nec \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(E) \biimp \nec E(\gamma)$}\RightLabel{$\biimp_E$}
\BIC{$\nec E(\gamma)$}
    \AXC{Axiom T} \dashedLine
    \UIC{$\nec E(\gamma) \imp E(\gamma)$} \RightLabel{$\imp_E$}
\BIC{$ E(\gamma)$} \dottedLine \RightLabel{D\ref{A:D3}}
\UIC{$ \all \varphi.[\ess{\varphi}{\gamma} \imp \nec \ex x.\varphi(x)] $}\RightLabel{$\all_E$}
\UIC{$ \ess{G}{\gamma} \imp \nec \ex x. G(x) $}\RightLabel{$\imp_E$}
\BIC{$\nec \ex x. G(x)$} \RightLabel{$\imp_I^1$}
\UIC{$\ex z. G(z) \imp \nec \ex x. G(x)$}
\end{prooftree}
\end{small}
\end{proof}

@@

\setcounter{lemma}{0}
\setcounter{theorem}{0}
\setcounter{corollary}{0}

@@

\begin{prooftree}
\AXC{$ $} \RightLabel{2}
\UIC{$\nec \neg \ex x.G_B(x) $} \RightLabel{$\nec_E$}
\UIC{$\omega: \neg \ex x.G_B(x) $} 
\AXC{$ $} \RightLabel{$1$}
\UIC{$\omega: G_B(x)$} \RightLabel{$\ex_I$}
\UIC{$\omega: \ex x.G_B(x) $} \RightLabel{$\imp_E$}
\BIC{$\omega:  \bot $}\RightLabel{$\imp_I^1$}
\UIC{$\omega:  \neg G_B(x) $} \RightLabel{$\imp_I$}
\UIC{$\omega:  G_B(x) \imp \neg G_B(x) $} \RightLabel{$\all_I$}
\UIC{$\omega:  \all x.(G_B(x) \imp \neg G_B(x)) $}  \RightLabel{$\nec_I$}
\UIC{$ \nec \all x.(G_B(x) \imp \neg G_B(x)) $}\dottedLine  \RightLabel{D\ref{B:D1}}
\UIC{$P(\neg G_B)$}
\AXC{Axiom 1} \dashedLine
\UIC{$\all \varphi. [P(\neg G_B) \imp \neg P(G_B)]$} \RightLabel{$\all_E$}
\UIC{$ P( \neg G_B(x)) \imp \neg P(G_B) $}  \RightLabel{$\imp_E$}
\BIC{$\neg P(G_B) $} 
\AXC{$ $} \RightLabel{3} 
\UIC{$P(G_B)$} 
\BIC{$\bot$} \RightLabel{$\imp_I^2$}
\UIC{$ \neg \nec \neg \ex x.G_B(x)$} \doubleLine %\RightLabel{definition of $\pos$}
\UIC{$ \pos \ex x.G_B(x)$}  \RightLabel{$\imp_I^3$}
\UIC{$ P(G_B) \imp \pos \ex x.G_B(x) $}

@@

\begin{corollary}
\label{B:C1}
Possibly, a God-like being exists:
$$
\pos \ex x. G(x)
$$
\end{corollary}
\begin{proof}

\begin{prooftree}
\AXC{$ $}\doubleLine
\UIC{$\nec \all x. [G_B(x) \imp G_B(x)] $}\RightLabel{ D\ref{B:D1}}\dottedLine
\UIC{$P_B(G_B) $}\RightLabel{ $\wedge_I$}
\AXC{Theorem \ref{T1}}
\UIC{$ P_B(G_B) \imp \pos \ex x.G_B(x) $} \RightLabel{ $\imp_E$}
\BIC{$\pos \ex x.G_B(x) $}
\end{prooftree}

@@

\begin{small}
\begin{prooftree}
\AXC{Axiom \ref{A:A5'}}\dashedLine
\UIC{$P_B (NE_B)$} \RightLabel{ D\ref{B:D1}}\dottedLine
\UIC{$\nec \all x. [G_B(x) \imp NE_B(x)]$}\RightLabel{ D\ref{B:D3}}\dottedLine
\UIC{$\nec \all x. [G_B(x) \imp \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$}  \RightLabel{ $\nec_E$}
\UIC{$\all x. [G_B(x) \imp \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$}  \RightLabel{ $\forall_E$}
\UIC{$ [G_B(x) \imp \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$}
\AXC{$ $} \RightLabel{ $2$}
\UIC{$G_B(x) $}  \RightLabel{ $\imp_E$}
\BIC{$\all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))$} \RightLabel{ $\forall_E$}
\UIC{$(MCP(G_B, x) \imp \nec \ex y.G_B(y))$}
\AXC{$ $}\dashedLine
\UIC{$MCP(G_B, x) $} \RightLabel{ $\imp_E$}
\BIC{$\nec \ex y.G_B(y)$}\RightLabel{ $\imp_I^2$}
\UIC{$G_B(x) \imp \nec \ex y.G_B(y)$}
\end{prooftree}
\end{small}
\end{proof}
